Discovering causal factors with back-to-back regression

### 2019-05-22 paper/paper.tex

Natural sciences are tasked to find, from a set of hypothetical factors, the minimal subset that suffices to reliably predict novel observations potentially in novel environments.
%
This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{n, q}$, that is a function of possible causes $X \in \mathbb{R}^{n, d}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{n, d}$: $Y=F(X,N)$. Of all considered causes (features of $X$), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{D}^{d, d}$ a binary diagonal matrix, which selects causal features of $X$.
%
Under the assumptions of centered and additive noise N, causal discovery thus consists in finding $E$ in

@@

    \node[] (labY) at (\posY, 2){Observations};
    \node[] (labX) at (\posX, 2){Factors};
    \node[] (labE) at (\posE, 2.3){Cause};
    \node[] (labE) at (\posE, 2){selection};
    \node[] (labN) at (\posN, 2){Noise};
    \node[] (labF) at (\posF, 2.3){Cause-effect};
    \node[] (labF) at (\posF, 2){mapping};

    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
    \node[] (reg1) at (5,-3.25) {$\hat{E} = \underbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}} $};
    \end{tikzpicture}
    \caption{In its linear form, back-to-back regression identifies the subset of factors that suffices to account for multidimensional observations by piping two successive regressions from $Y$ to $X$ and from $X$ to $\hat X$}
    \label{fig:}
\end{figure}

@@

This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance).
%
In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
%
If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximizing the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
%
If diagonal elements of $\hat H$ are normally distributed, K follows a chi-square distribution with one degree of freedom.
%
At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}

>>>>

Natural sciences are tasked to find, from a set of hypothetical factors, the minimal subset that suffices to reliably predict novel observations potentially in novel environments.
%
This problem can be formalized as follow: empirical studies consist in measuring, via an apparatus $F$, a signal $Y \in \mathbb{R}^{i, l}$, that is a function of possible causes $X \in \mathbb{R}^{i, j}$, and other unknown causes that can be understood as noise $N \in \mathbb{R}^{i, k}$: $Y=F(X,N)$. Of all considered causes (features of $X$), only a subset accounts for $Y$, let it be represented as $E \cdot X$, with $E \in \mathbb{R}^{j, j}$ a binary diagonal matrix, which selects causal features of $X$.
%
Under the assumptions of centered and additive noise N, causal discovery thus consists in finding $E$ in

@@

    \node[] (labY) at (\posY, 2){observed effects};
    \node[] (labX) at (\posX, 2.3){world};
    \node[] (labX) at (\posX, 2){features};
    \node[] (labE) at (\posE, 2.3){cause};
    \node[] (labE) at (\posE, 2){selection};
    \node[] (labN) at (\posN, 2){noise};
    \node[] (labF) at (\posF, 2.3){cause-effect};
    \node[] (labF) at (\posF, 2){mapping};

    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
    \node[] (reg1) at (5,-3.25) {$\hat{E} = \overbrace{(X_2^\top X_2 + \lambda_X)^{-1} X_2^\top Y_2\underbrace{(Y_1^\top Y_1 + \lambda_Y)^{-1} Y_1^\top X_1}_{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}^{\text{2) regression from } X \text{ to } \hat{X}} $};
    \end{tikzpicture}
    \caption{Summary of our method blah blah blah}
    \label{fig:}
\end{figure}

@@

This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance).
%
In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
%
If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximising the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
%
If diagonal elements of $\hat H$ are normally distributed, K follows a chi-square distribution with one degree of freedom.
%
At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}
